Survival in chessland

### 2019-03-02 chess/papers/survival.tex
Submitted version

  Since this is one of the shortest possible routes to a draw, I
  hereby dub this line the ``Fool's Draw,'' by analogy with the Fool's
  Mate. In this line all pieces survive, which is anyway humane and
  also advantageous in the case that you are someone is simultaneously
  being one of the chesspieces to the death!

@@

\begin{itemize}
\item Black and white are probably not significantly different. That
  is, the a1 and a8 rooks probably have about the same survival
  chances (it's known that white has a slight statistical
  advantage~\cite{chesstego} but it is probably only around 1\%). So
  these guesses will be written about white's pieces.
\item The d2 and e2 pawns are very active in common openings, and
  are frequently captured as part of those openings. I think they
  are the least likely to survive overall.
\item Bishops and knights are often involved in the opening and
  midgame, and often exchanged nonchalantly. I think they all have
  relatively low survival chances.
\item Although the queen is very valuable, a queen exchange is often
  forced for games that enter the endgame.
\item Rooks tend to be late-game pieces, because they are difficult
  to get out of their corners (and at most one can be activated
  by the fastest method, castling) and are relatively valuable.
\item This leaves the non-central pawns. These are the hardest to
  predict, and they are hard to think about (at least for me) because
  when e.g. the a2 pawn recaptures the b3 pawn that it supported, I
  just think of this as the b3 pawn. Of these pawns, b2 and g2 are
  somewhat weak because they are undefended once the bishop is
  developed (cf. the famous ``poison pawn'' at b2). On the other hand,
  in the fianchetto configuration, this pawn is very strong and often
  survives the entire game without leaving the third rank. Since pawn
  chains usually progress towards the middle of the board, the a2 pawn
  is more likely to be supporting than supported. This both leaves it
  weak to capture, but prone to recapturing. Outside pawns block one's
  own rook, although for this same reason they often clear the file by
  capturing (and so survive). They are also commonly used to push into
  a well-defended king's territory (e.g. in the fianchetto); kingside
  castling is more common, so this means that the h pawns are often
  lost to this fate.
\end{itemize}

The final ranking that I committed to, from most surviving to most dead:
\pawn f, \pawn c, \pawn g, \pawn a, \pawn h, \pawn b, \rook h, \rook a,
\king, \queen, \bishop f, \knight g, \pawn e, \pawn d.

@@

The basic survival chances appear in Figure~\ref{fig:piece-survival}.
Indeed, many pieces are more likely to survive than the kings. Even as
black, the extremal pawns (\pawn a and \pawn h) have over a 70\%
survival rate. Across the board, the survival chances for a white
piece and its black twin are similar, usually with a small edge to
white. Notable exceptions are the \knight g (the overall most doomed
piece), and both white bishops, which die more than their
Schwarzdoppelgangers. The \pawn e is vastly more dead than \Pawn e.
Note somewhat satisfyingly that the \pawn c has the highest variance;
this was the most controversial among the drinking buddies
(Section~\ref{sec:guesses}). Note that \pawn c is the sacrificed
pawn in the popular Queen's Gambit (1. d4 d5 2. c4), where accepting
and declining the pawn are both popular and sound responses. This may
be a good example of a piece that has substantially different survival
rates in different opening preferences. Since the variance is
otherwise extremely low, I only report means for the remainder of the
paper.

@@

% XXX using this a way to play the game

Other than the bishops---which have no legal way to reach half of the
squares---every piece ends on every square in at least a thousand
games. So we have enough samples to have reasonable confidence in our
statistics, even for the most unlikely odysseys. The least mobile
pieces are the pawns, who can technically reach any square by
promoting, but are usually confined to cones emanating from their
start squares. The overall rarest fate is for \pawn f2 to die on the
a7 square, which only happened 1,244 times (however, it survived on
this square 31,438 times). This square is actually reachable without
promoting, but it would need to capture 5 times in order to get
there, which seems quite unlikely! There may even be a hidden
achievement for reaching this square this way!
% 12 times for titled players (in new dataset too), although there are
% a few things that are even rarer, like dying on a8.
% TODO: Perhaps could find a game where this happened?!
Aside from pawns, the weirdest fate is for \knight g to die on h1,
which happened 47,307 times. Corners are of course fairly garbage for
the knight, although it is twice as likely to survive on this square.

% There may even be hidden achievements for reaching these squares!

That said, there are characteristic patterns for each of the pieces,
which make sense given their starting positions and patterns of
movement. You could probably guess the piece just by looking at one of
the heat maps below, although---spoiler alert---the piece is just
listed right there and they are in order. Two independent things are
communicated in these graphics: The chance that the piece ends the
game on some square (alive or dead), and its survival chances there.
In each map, a darker background color indicates that the piece ends
on that square more often. The shade is based on the rank (64/64 black
is most common, 63/64 black is next most common, etc.) rather than the
absolute probability, since otherwise the graphic looks boring. The
number on the square is the percentage survival of the piece when it
is last seen (either being captured or surviving to the end of the
game) on that square. The four squares with the highest survival rates
are underlined for your convenience.

@@

@@

\begin{figure}
  \includegraphics[width=\linewidth]{hypotheses}
  \caption{
    Piece rankings (from most surviving to most dead); either a
    human or hypothesis or the actual results across all games. 
    The {\bf actual} column appears multiple times so that each human
    gets a chance to be adjacent to it; this makes for the easiest
    visual comparison. Lines connect the same piece in adjacent columns,
    and are darker if the pairs have more different ranks.
  } \label{fig:buddies}
\end{figure}

Like all good scientific research, I explicitly compare the actual
results to the hypotheses gathered before the experiment; this is a
hygenic and humbling exercise. Figure~\ref{fig:buddies} compares the
ranking across all games (slice {\bf Actual}) to the author author
(slice {\bf Tom}) and his drinking buddies (slices {\bf Ben}, {\bf
  Jim}, {\bf David}, {\bf William}). There are a number of different
reasonable ways to measure the accuracy of this type of position; a
very simple one is the sum of the absolute differences in rank for
each piece (e.g. if in one ranking \king is \#3, and in the other \#5,
then this contributes 2 to the total error). By this metric, Ben has
the best prediction (98 error), followed by David (138) and Tom (148).
Tom and David had the most similar predictions (116) and Jim and
William the most different (312). The expected error between two
completely random permutations is about 341, so all of these guesses
are significantly better than chance. Note in the actual ranking, many
pieces have very similar survival probabilities, and many guesses are
ambivalent about groups of pieces. Weighting each rank difference
equally is therefore an oversimplification. It would have been better
to ask each participant to give probabilities, as David did; this
would give us more sensitive error metrics and more opportunities to
spend the afternoon making visualizations.

@@

@@

\begin{figure}
  \includegraphics[width=\linewidth]{slices}
  \caption{
    Ranks and survival probabilities for different subsets of games.
    The same piece in adjacent rows is connected to highlight
    differences in the ranking, as before. The time formats all
    exhibit similar ranking with only small perturbations. Games
    including a titled player (rightmost column) are the most
    different, although we have far fewer samples in this set,
    so variance becomes significant.
  } \label{fig:slices}
\end{figure}

The survival probabilities differ depending on the conditions of the
game; Figure~\ref{fig:slices} compares some of those slices. Here the
{\bf All} slice is the same as the {\bf Actual} slice in
Figure~\ref{fig:buddies}, and consists of all acceptable games in the
database.

The {\bf titled} slice includes only games where at least one of the
players has an official title (Grandmaster, International Master, FIDE
Master, etc.\footnote{Lichess used to award the LM ``Lichess Master''
  title to notable players on the site; this title is excluded from
  the sample.}). These games have high-quality play, but far fewer
samples (``only'' 3.4 million). This set exhibits significant
variance; for example, the survival rate of \pawn c ranges from
$38$--$46\%$. This is both because of the small sample size and the
bucketization by player name; there are few enough titled players that
an individual's preference in openings and style of play moves the
values of their entire bucket. I caution against reading too much into
this column. It seems we can at least conclude that these games tend
to be much bloodier (Kings aside, survival rates are lower across the
board); top players are less likely to fall for traps early in the
game, perhaps more willing to sacrifice material, and more likely to
play into endgames where almost every piece is exchanged. If being a
chesspiece to the death, you do {\em not} want to have a Grandmaster
playing the game!

@@

On the other hand, the other slices all have enough samples that the
variance is minimal. These slices, {\bf bullet} (151,261,707 games),
{\bf blitz} (236,050,938 games), {\bf rapid} (98,606,558) and {\bf
  classical} (13,886,352) are different time control formats. Games on
Lichess are played with a starting clock (per side) and an increment
added to the clock after each player's move. The game is classified
according to the estimated total time: The starting time plus $40 \times$
the increment (with the idea that an average game has 40 moves per
side); this is the same formula that Lichess uses. A bullet game is
when the total time per side is between 30~seconds and 3~minutes;
blitz is between 3 and 8~minutes; rapid is between 8 and 25~minutes;
classical is any more than this (including untimed games). Each of
these slices has enough samples that the variance is very low.
Here we see that the ranking is rather stable across the range of
time formats, which was not what I expected. This should increase
our confidence that the results are really inherent to chess, not
just the particulars of this data set.

>>>>

\clearpage

  Since this is one of the shortest possible routes to a draw, I
  hereby dub this line the ``Fool's Draw,'' by analogy with the Fool's
  Mate. In this line all pieces survive, which is anyway humane and
  also advantageous in the case that you or someone is simultaneously
  being one of the chesspieces to the death!

@@

\begin{itemize}
\item Black and white are probably not substantially different. That
  is, the a1 and a8 rooks probably have about the same survival
  chances (it's known that white has a slight statistical
  advantage~\cite{chesstego} but it is probably only around 1\%). So
  these guesses will be written about white's pieces.
\item The d2 and e2 pawns are very active in common openings, and
  are frequently captured as part of those openings. I think they
  are the least likely to survive overall.
\item Bishops and knights are often involved in the opening and
  midgame, and often exchanged nonchalantly. I think they all have
  relatively low survival chances.
\item Although the queen is very valuable, a queen exchange is often
  forced for games that enter the endgame.
\item Rooks tend to be late-game pieces, because they are difficult
  to get out of their corners (and at most one can be activated
  by the fastest method, castling) and are relatively valuable.
\item This leaves the non-central pawns. These are the hardest to
  predict, and they are hard to think about (at least for me) because
  when e.g. the a2 pawn recaptures the b3 pawn that it supported, I
  just think of this as the b3 pawn. Of these pawns, b2 and g2 are
  somewhat weak because they are undefended once the bishop is
  developed (cf. the famous ``poison pawn'' at b2). On the other hand,
  in the fianchetto configuration, this pawn is very strong and often
  survives the entire game without leaving the third rank. Since pawn
  chains usually progress towards the middle of the board, the a2 pawn
  is more likely to be supporting than supported. This both leaves it
  weak to capture, but prone to recapturing. Outside pawns block one's
  own rook, although for this same reason they often clear the file by
  capturing (and so survive). They are also commonly used to push into
  a well-defended king's territory (e.g. in the fianchetto); kingside
  castling is more common, so this means that the h pawns are often
  lost to this fate.
\end{itemize}

The final ranking that I predict, from most surviving to most dead:
\pawn f, \pawn c, \pawn g, \pawn a, \pawn h, \pawn b, \rook h, \rook a,
\king, \queen, \bishop f, \knight g, \pawn e, \pawn d.

@@

The basic survival chances appear in Figure~\ref{fig:piece-survival}.
Indeed, many pieces are more likely to survive than the kings. Even as
black, the extremal pawns (\pawn a and \pawn h) have over a 70\%
survival rate. Across the board, the survival chances for a white
piece and its black twin are similar, usually with a small edge to
white. Notable exceptions are the \knight g (the overall most doomed
piece), and both white bishops, which die more than their
Schwarzdoppelgangers. The \pawn e is vastly more dead than \Pawn e.
Note somewhat satisfyingly that the \pawn c has the highest variance;
this was the most controversial among the drinking buddies
(Section~\ref{sec:guesses}). Note that \pawn c is the sacrificed
pawn in the popular Queen's Gambit (1. d4 d5 2. c4), where accepting
and declining the pawn are both common and sound responses. This may
be a good example of a piece that has substantially different survival
rates in different opening preferences. Since the variance is
otherwise extremely low, I only report means for the remainder of the
paper.

@@

Other than the bishops---which have no legal way to reach half of the
squares---every piece ends on every square in at least a thousand
games. So we have enough samples to have reasonable confidence in our
statistics, even for the most unlikely odysseys. The least mobile
pieces are the pawns, who can technically reach any square by
promoting, but are usually confined to cones emanating from their
start squares. The overall rarest fate is for \pawn f2 to die on the
a7 square, which only happened 1,244 times (however, it survived on
this square 31,438 times). This square is actually reachable without
promoting, but it would need to capture 5 times in order to get
there, which seems quite unlikely! There may even be a hidden
achievement for reaching this square this way!
% 12 times for titled players (in new dataset too), although there are
% a few things that are even rarer, like dying on a8.
% TODO: Perhaps could find a game where this happened?!
Aside from pawns, the weirdest fate is for \knight g to die on h1,
which happened 47,307 times. Corners are of course garbage for the
knight, although it is twice as likely to survive on this square.

There are characteristic patterns for each of the pieces, which make
sense given their starting positions movement rules. You could
probably guess the piece just by looking at one of the heat maps
below, although---spoiler alert---the piece is just listed right there
and they are in order. Two independent things are communicated in
these graphics: The chance that the piece ends the game on some square
(alive or dead), and its survival chances there. In each map, a darker
background color indicates that the piece ends on that square more
often. The shade is based on the rank (64/64 black is most common,
63/64 black is next most common, etc.) rather than the absolute
probability, since otherwise the graphic looks boring. The number on
the square is the percentage survival of the piece when it is last
seen (either being captured or surviving to the end of the game) on
that square. The four squares with the highest survival rates are
underlined for your convenience.

@@

Aside from admiring the groovy pictures, the data can also be used as
the basis of an algorithm for playing chess. There are several things
to try; we could move pieces towards squares where they are more
likely to survive than die (keeping pieces alive is good), or just
towards squares where they are more likely to be positioned at the end
of the game (our moves are more likely to resemble real chess moves
because they put pieces in their proper places). Alas, it turns out
that these are bad approaches to chess, both because they are boring
(most pieces are actually most comfortable in their starting
positions) and because they perform badly against even weak
opponents~\cite{eloworld}.

@@

Like all good scientific research, I explicitly compare the actual
results to the hypotheses gathered before the experiment; this is an
hygenic and humbling exercise. Figure~\ref{fig:buddies} compares the
ranking across all games (slice {\bf Actual}) to the author author
(slice {\bf Tom}) and his drinking buddies (slices {\bf Ben}, {\bf
  Jim}, {\bf David}, {\bf William}). There are a number of different
reasonable ways to measure the accuracy of this type of position; a
very simple one is the sum of the absolute differences in rank for
each piece (e.g. if in one ranking \king is \#3, and in the other \#5,
then this contributes 2 to the total error). By this metric, Ben has
the best prediction (98 error), followed by David (138) and Tom (148).
Tom and David had the most similar predictions (116) and Jim and
William the most different (312). The expected error between two
completely random permutations is about 341, so all of these guesses
are significantly better than chance. Note in the actual ranking, many
pieces have very similar survival probabilities, and many guesses are
ambivalent about groups of pieces. Weighting each rank difference
equally is therefore an oversimplification. It would have been better
to ask each participant to give probabilities, as David did; this
would give us more sensitive error metrics and more opportunities to
spend the afternoon making visualizations.

@@

\begin{figure}[t]
  \includegraphics[width=\linewidth]{hypotheses}
  \caption{
    Piece rankings (from most surviving to most dead); either a
    human or hypothesis or the actual results across all games. 
    The {\bf actual} column appears multiple times so that each human
    gets a chance to be adjacent to it; this makes for the easiest
    visual comparison. Lines connect the same piece in adjacent columns,
    and are darker if the pairs have more different ranks.
  } \label{fig:buddies}
\end{figure}

@@

\begin{figure}[t]
  \includegraphics[width=\linewidth]{slices}
  \caption{
    Ranks and survival probabilities for different subsets of games.
    The same piece in adjacent rows is connected to highlight
    differences in the ranking, as before. The time formats all
    exhibit similar ranking with only small perturbations. Games
    including a titled player (rightmost column) are the most
    different, although we have far fewer samples in this set,
    so variance becomes significant.
  } \label{fig:slices}
\end{figure}

The survival probabilities differ depending on the conditions of the
game; Figure~\ref{fig:slices} compares some of those slices. Here the
{\bf All} slice is the same as the {\bf Actual} column in
Figure~\ref{fig:buddies}, and consists of all acceptable games in the
database.

The {\bf Titled} slice includes only games where at least one of the
players has an official title (Grandmaster, International Master, FIDE
Master, etc.\footnote{Lichess used to award the LM ``Lichess Master''
  title to notable players on the site; this title is excluded from
  the sample.}). These games have high-quality play, but far fewer
samples (``only'' 3.4 million). This set exhibits significant
variance; for example, the survival rate of \pawn c ranges from
$38$--$46\%$. This is both because of the small sample size and the
bucketization by player name; there are few enough titled players that
an individual's preference in openings and style of play changes the
values of their entire bucket. I caution against reading too much into
this column. It seems we can at least conclude that these games tend
to be much bloodier (Kings aside, survival rates are lower across the
board); top players are less likely to fall for traps early in the
game, perhaps more willing to sacrifice material, and more likely to
play into endgames where almost every piece is exchanged. If being a
chesspiece to the death, you do {\em not} want to have a Grandmaster
playing the game!

@@

On the other hand, the other slices all have enough samples that the
variance is minimal. These slices, {\bf Bullet} (151,261,707 games),
{\bf Blitz} (236,050,938 games), {\bf Rapid} (98,606,558) and {\bf
  Classical} (13,886,352) are different time control formats. Games on
Lichess are played with a starting clock (per side) and an increment
added to the clock after each player's move. The game is classified
according to the estimated total time: The starting time plus $40 \times$
the increment (with the idea that an average game has 40 moves per
side); this is the same formula that Lichess uses. A bullet game is
when the total time per side is between 30~seconds and 3~minutes;
blitz is between 3 and 8~minutes; rapid is between 8 and 25~minutes;
classical is any more than this (including untimed games). Each of
these slices has enough samples that the variance is very low.
Here we see that the ranking is rather stable across the range of
time formats, which was not what I expected. This should increase
our confidence that the results are really inherent to chess, not
just the particulars of this data set.
